```{r, include=FALSE}
knitr::opts_chunk$set(comment = "")
```

# Plotting in R

There are two major sets of tools for creating plots in R, namely the **base** and **ggplot2** facilities.  Base graphics come with all R installations, while ggplot2 is a stand-alone package.  Note that other plotting facilities do exist (notably **lattice**), but base and ggplot2 are by far the most popular.  

### The dataset
For the following examples, we will use a dataset taken from Fair (1977) in **Econometrica**, entitled "A note on the computation of the tobit estimator".  Improbably, this dataset contains information on extramarital affairs. 
```{r}
library(Ecdat)
data(Fair)
ema<-Fair
head(ema)
str(ema)
library(plyr)
ema<-ddply(.data = ema, .variables = "age", .fun = transform, avg.nbaffairs=mean(nbaffairs, na.rm=T))
```

### base

* *Minimal* call takes the following form

```{r, eval=F}
plot(x=)
```
```{r}
# Note that when asked to plot a single vector, R will assume the index positions of each vector element are the implied horizontal dimension
plot(x = ema$age) 
```

* *Basic* call takes the following form

```{r, eval=F}
plot(x=, y=)
```
```{r}
plot(x = ema$age, y = ema$avg.nbaffairs)
```

### base/type (scatter, line, both)

* The "type" argument accepts the following character indicators
* "p" -- point/scatter plots (default plotting behavior)
```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="p")
```
* "l" -- line graphs
```{r, fig.cap=" "}
# Note that "line" does not create a smoothing line, just connected points
plot(x=ema$age, y=ema$avg.nbaffairs, type="l") 
```
* "b" -- both line and point plots
```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="b") 
```

### base/type (histograms, density plots)

* Certain plot types require different calls outside of the "type" argument
* Ex) Histograms
```{r, fig.cap=" "}
hist(x=ema$age)
hist(x=ema$age, breaks=5)
hist(x=ema$age, breaks=10)
```
* Ex) Density plots
```{r, fig.cap=" "}
age.density<-density(x=ema$age, na.rm=T) # Create a density object (NOTE: be sure to remove missing values)
class(age.density) # Check the class of the object
age.density # View the contents of the object
plot(x=age.density) # Plot the density object
plot(x=density(x=ema$age, bw=.5, na.rm=T)) # Plot the density object, bandwidth of 0.5
plot(x=density(x=ema$age, bw=2, na.rm=T)) # Plot the density object, bandwidth of 2
plot(x=density(x=ema$age, bw=6, na.rm=T)) # Plot the density object, bandwidth of 6
```

### base/options [labeling]

* Basic call with popular labeling arguments
```{r, eval=F}
plot(x=, y=, type="", xlab="", ylab="", main="") 
```
* From the previous example
```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", xlab="Age (Years)", ylab="Number of Affairs (per 12 months)", main="Annual Extramarital Affairs by Age") # Add labels for axes and overall plot
```

### base/options [axis + size scaling]

* Basic call with popular scaling arguments
```{r, eval=F}
plot(x=, y=, type="", xlim=, ylim=, cex=)
```
* From the previous example

```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="p") # Create a basic plot
# Limit the ages (x-axis) to between 20 and 40
plot(x=ema$age, y=ema$avg.nbaffairs, type="p", xlim = c(20,40)) 
# Limit the ages (x-axis) to between 20 and 40, increase point size to 2
plot(x=ema$age, y=ema$avg.nbaffairs, type="p", xlim = c(20,40), cex=2) 
# Limit the ages (x-axis) to between 20 and 40, decrease point size to 0.5
plot(x=ema$age, y=ema$avg.nbaffairs, type="p", xlim = c(20,40), cex=0.5)  
```

### base/options [graphical parameters]

* Basic call with popular scaling arguments
```{r, eval=F}
plot(x=, y=, type="", col="", pch=, lty=, lwd=)
```
* Colors
```{r}
colors() # View all elements of the color vector
colors()[179] # View specific element of the color vector
```
Another option: [R Color Infographic](http://research.stowers-institute.org/efg/R/Color/Chart/ColorsChart1.jpg)
```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", col=colors()[145]) # or col="gold3"
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", col="seagreen4") # or col=colors()[578]
```

* Point Styles and Widths

[A Good Reference](http://www.endmemo.com/program/R/pic/pchsymbols.png)

```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="p", pch=3) # Change point style to crosses
plot(x=ema$age, y=ema$avg.nbaffairs, type="p", pch=15) # Change point style to filled squares
# Change point style to filled squares and increase point size to 3
plot(x=ema$age, y=ema$avg.nbaffairs, type="p", pch=15, cex=3) 
plot(x=ema$age, y=ema$avg.nbaffairs, pch="w") # Change point style to "w"
# Change point style to "$" and increase point size to 2
plot(x=ema$age, y=ema$avg.nbaffairs, type="p", pch="$", cex=2) 
```

* Line Styles and Widths
```{r, fig.cap=" "}
# Line plot with solid line
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=1)
# Line plot with medium dashed line
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=2)
# Line plot with short dashed line
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=3)
# Change line width to 2
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=3, lwd=2)
# Change line width to 5
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=3, lwd=5)
# Change line width to 10 and use dash-dot
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=4, lwd=10)
```

### base/options [annotations/reference lines/legends]

* Text 
```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=4, lwd=10)
text(x=25, y=3, labels="Sharp Drop", cex = .75)
```

* Reference Lines
```{r, fig.cap=" "}
plot(x=ema$age, y=ema$avg.nbaffairs, type="l", lty=4, lwd=10)
abline(v=42.5, h=2.25, lty=2)
```

### ggplot2

The general call for ggplot2 takes on one of three forms:
  
```{r, eval=FALSE}
# 1) Aesthetics/options are specified the initial ggplot() layer and will be used by all other layers
ggplot(data=, aes(x=,y=), color=, size=,)+geom_xxxx()+geom_yyyy()

# 2) Aesthetics/options are not specified the initial ggplot() layer, but in other layers
ggplot()+geom_xxxx(data=, aes(x=,y=), color=, size=,)+geom_yyyy(data=, aes(x=,y=), color=, size=,)

# 3) Aesthetics/options are specified partially in the ggplot() layers and partially in other layers  
ggplot(data=, aes(x=,y=), color=, size=,)+geom_xxxx()+geom_yyyy(data=, aes(x=,y=), color=, size=,)
```

In general (but with some exceptions), aesthetics/options specified in the ```ggplot()``` layer will override aesthetics/options specified in other layers and will be the source for required argument values not present in those layers. 

### Density plots

```{r, fig.cap=" ", warning=FALSE}
library(ggplot2)
ggplot(data=ema, aes(x=age))+geom_density()
ggplot()+geom_density(data=ema, aes(x=age))
ggplot(data=ema)+geom_density(aes(x=age))
ggplot(data=ema, aes(x=age))+geom_line(stat="density") # Trick for removing coverage line
ggplot(data=ema, aes(x=age))+geom_line(stat="density")+theme_bw() # Trick for removing background shading
```

### X-Y scatter plots

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_point()+theme_bw()
```

### X-Y line plots

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line()+theme_bw()
```

### bar plots 
```{r, fig.cap=" ", warning=FALSE, message=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_bar(stat="identity")+theme_bw()
```

### box plots 
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=sex,y = nbaffairs))+geom_boxplot()+theme_bw()
ggplot(data=ema, aes(x=child,y = nbaffairs))+geom_boxplot()+theme_bw()
ggplot(data=ema, aes(x=factor(religious),y = nbaffairs))+geom_boxplot()+theme_bw()
ggplot(data=ema, aes(x=factor(rate),y = nbaffairs))+geom_boxplot()+theme_bw()
```

### "trellis" plots 
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=factor(rate),y = nbaffairs))+geom_boxplot()+theme_bw()+facet_wrap(facets = ~sex)
ggplot(data=ema, aes(x=sex,y = nbaffairs))+geom_boxplot()+theme_bw()+facet_wrap(facets = ~rate+child)
```

### options [labeling]

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line()+theme_bw()
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line()+theme_bw()+
  xlab("Age (Years)")+ylab("Number of Affairs (per 12 months)")+
  ggtitle("Annual Extramarital Affairs by Age")
```

### options [axis + size scaling]

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_point()+theme_bw()+
  xlim(20,40)
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_point(size=.5)+theme_bw()+
  xlim(20,40)
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_point(size=10)+theme_bw()+
  xlim(20,40)
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line()+theme_bw()+
  xlim(20,40)
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(size=.1)+theme_bw()+
  xlim(20,40)
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(size=3)+theme_bw()+
  xlim(20,40)
```

### options [graphical parameters]

* Colors
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(color="red")+theme_bw()
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(color="seagreen4")+theme_bw()
```

* Point Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_point(shape=3)+theme_bw()
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_point(shape=15)+theme_bw()
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_point(shape="@", size=10)+theme_bw()
```

* Line Styles and Widths
```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(linetype=2)+theme_bw()
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(linetype=4)+theme_bw()
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(linetype=2, size=.5)+theme_bw()
ggplot(data=ema, aes(x=age,y = avg.nbaffairs))+geom_line(linetype=4, size=10)+theme_bw()
```

### Anatomy of aes()

ggplot2 tries to represent variation in data by making use of various "aesthetics".  This general approach is based on *The Grammar of Graphics*, which says that visualizing data requires four elements:

1) One or more **statistics** conveying information about the data (identities, means, medians, etc.)

2) A **coordinate system** that differentiates between the intersections of statistics (at most two for ggplot, three for lattice)

3) **Geometries** that differentiate between off-coordinate variation in *kind*
  
4) **Scales** that differentiate between off-coordinate variation in *degree*

All of these elements can be manipulated to display variation in the data via the ```aes()``` function.  
```{r, eval=FALSE}
ggplot(data=, aes(x=, y=, color=, linetype=, shape=, size=))
```

Normally, specifying options like ```color="red"``` or ```size=10``` for a given layer results in its contents being red and quite large.  Inside the ```aes()``` function, however, these arguments are given entire variables whose values will then be displayed using different realizations of that aesthetic.

```{r, fig.cap=" ", warning=FALSE}
# Compute average number of affairs by sex
ema<-ddply(.data = ema, .variables = c("age", "sex"), .fun = transform, 
           avg.nbaffairs.age.sex=mean(nbaffairs, na.rm=T))
# Differences in kind using color
ggplot(data=ema, aes(x=age, y=avg.nbaffairs.age.sex))+geom_line()+theme_bw()
ggplot(data=ema, aes(x=age, y=avg.nbaffairs.age.sex, color=sex))+geom_line()+theme_bw()
```

```{r, fig.cap=" ", warning=FALSE}
ggplot(data=ema, aes(x=age, y=avg.nbaffairs.age.sex, linetype=sex))+geom_line()+theme_bw()
ggplot(data=ema, aes(x=age, y=avg.nbaffairs.age.sex, shape=sex))+geom_point()+theme_bw()
ggplot(data=ema, aes(x=age, y=religious, color=nbaffairs))+geom_jitter()+theme_bw() # Note the use of the jitter layer
ggplot(data=ema, aes(x=age, y=religious, size=nbaffairs))+geom_jitter()+theme_bw() # Note the use of the jitter layer
```

### Fitted lines and curves 
```{r, fig.cap=" ", warning=FALSE, message=FALSE}
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth(method="lm")
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth(method="loess")
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth(method="loess", span=.5)
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth(method="loess", span=.99)
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth()
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth(se=F)
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth(se=F, color="red")
ggplot(data=ema, aes(x=age, y=nbaffairs))+geom_point()+theme_bw()+geom_smooth(aes(color=sex, fill=sex))
```

### Exporting

Two basic image types

1) **Raster/Bitmap** (.png, .jpeg)

Every pixel of a plot contains its own separate coding; not so great if you want to resize the image

```{r, eval=FALSE}
jpeg(filename="example.png", width=, height=)
plot(x,y)
dev.off()
```

2) **Vector** (.pdf, .ps)

Every element of a plot is encoded with a function that gives its coding conditional on several factors; great for resizing

```{r, eval=FALSE}
pdf(filename="example.pdf", width=, height=)
plot(x,y)
dev.off()
```

### Exporting with ggplot

```{r, eval=FALSE}
# Assume we saved our plot is an object called example.plot
ggsave(filename="example.pdf", plot=example.plot, scale=, width=, height=)
```
