---
title: "Introduction to R"
author: "Rochelle Terman"
date: "August 12, 2015"
output: html_document
---

# Introduction to R

R is a versatile, open source programming/scripting language that's useful both for statistics but also data science. Inspired by the programming language [`S`][S].

* Open source software under [GPL][].
* Superior (if not just comparable) to commercial alternatives. R has over 5,000 user contributed packages at this time. It's widely used both in academia and industry.
* Available on all platforms.
* As a result, if you do your analysis in R, anyone can easily replicate it.
* Not just for statistics, but also general purpose programming.
* Is object oriented and functional.
* Large and growing community of peers.

[S]: http://stat.bell-labs.com/S/
[GPL]: https://www.gnu.org/licenses/gpl.html

### R Environment

**Viewing objects in your global environment and how to clean them up**

List objects in your current environment

```{r, eval = FALSE}
ls()
```

remove objects from your current environment

```{r, eval = TRUE, echo = TRUE}
x <- 5
rm(x)
#print(x)
```

remove all objects from your current environment

```{r, eval = FALSE}
rm(list = ls())
```

*Notice that we have nested one function inside another.*

**Package management**

* `install.packages("package-name")` will download a package from one of the CRAN mirrors assuming that a binary is available for your operating system. If you have not set a preferred CRAN mirror in your `options()`, then a menu will pop up asking you to choose a location.
* `library("package-name")` will load a package so you can use it. It is required at the beginning of each R session.

```{r, eval = FALSE}
library(stats)
```

### Basic Syntax

**Comments**

Use `#` signs to comment. Comment liberally in your R scripts. Anything to the right of a `#` is ignored by R. There is no docstring, or equivalent to `"""` in R

**Variable Names**

Variable names can only contain letters, numbers, the underscore character, and (unlike Python) the period character. Whereas an object name like ```myobject.thing``` would point to the subclass ```thing``` of ```myobject```, R only understands the object to be freestanding.  Put another way, R interprets ```myobject.thing``` the same way that Python would interpret ```myobject_thing```.

**Assignment operator**

`<- ` is the assignment operator. Assigns values on the right to objects on the left. Mostly similar to `=` but not always. Learn to use `<-` as it is good programming practice. Using `=` in place of `<-` can lead to issues down the line.

```{r}
experiment<-"current vs. voltage"
```

**Printing**
In R, the contents of an object can be printed by either simply executing the the object name or calling the ```print()``` function.  Note that, unlike in Python 2, ```print()``` must have object names passed directly to it as arguments inside parentheses. (Note: This changes in Python 3)

```{r, error=FALSE}
# Executing the object name
experiment
# Printing the object R-style
print(experiment)
```

**Help**

* `?` + object opens a help page for that specific object
* `??` + object searches help pages containing the name of the object

```{r, eval = FALSE}
?mean
??mean
```

#### Bonus: More about R

From [Hadley Wickham](http://adv-r.had.co.nz/Introduction.html):

If you are new to R, you might wonder what makes learning such a quirky language worthwhile. To me, some of the best features are:

* It's free, open source, and available on every major platform. As a result, if 
  you do your analysis in R, anyone can easily replicate it.

* A massive set of packages for statistical modelling, machine learning,
  visualisation, and importing and manipulating data. Whatever model or
  graphic you're trying to do, chances are that someone has already tried
  to do it. At a minimum, you can learn from their efforts.

* Cutting edge tools. Researchers in statistics and machine learning will often
  publish an R package to accompany their articles. This means immediate
  access to the very latest statistical techniques and implementations.

* Deep-seated language support for data analysis. This includes features
  likes missing values, data frames, and subsetting.

* A fantastic community. It is easy to get help from experts on the
  [R-help mailing list](https://stat.ethz.ch/mailman/listinfo/r-help),
  [stackoverflow](http://stackoverflow.com/questions/tagged/r),
  or subject-specific mailing lists like
  [R-SIG-mixed-models](https://stat.ethz.ch/mailman/listinfo/r-sig-mixed-models)
  or [ggplot2](https://groups.google.com/forum/#!forum/ggplot2). You
  can also connect with other R learners via
  [twitter](https://twitter.com/search?q=%23rstats),
  [linkedin](http://www.linkedin.com/groups/R-Project-Statistical-Computing-77616),
  and through many local
  [user groups](http://blog.revolutionanalytics.com/local-r-groups.html).

* Powerful tools for communicating your results. R packages make it easy to
  produce html or pdf [reports](http://yihui.name/knitr/), or create
  [interactive websites](http://www.rstudio.com/shiny/).

* A strong foundation in functional programming. The ideas of functional
  programming are well suited to solving many of the challenges of data
  analysis. R provides a powerful and flexible toolkit which allows
  you to write concise yet descriptive code.

* An [IDE](http://www.rstudio.com/ide/) tailored to the needs of interactive
  data analysis and statistical programming.

* Powerful metaprogramming facilities. R is not just a programming language, it
  is also an environment for interactive data analysis. Its metaprogramming
  capabilities allow you to write magically succinct and concise functions and
  provide an excellent environment for designing domain-specific languages.

* Designed to connect to high-performance programming languages like C,
  Fortran, and C++.

Of course, R is not perfect. R's biggest challenge is that most R users are not programmers. This means that:

* Much of the R code you'll see in the wild is written in haste to solve
  a pressing problem. As a result, code is not very elegant, fast, or easy to
  understand. Most users do not revise their code to address these shortcomings.

* Compared to other programming languages, the R community tends to be more
  focussed on results instead of processes. Knowledge of software engineering
  best practices is patchy: for instance, not enough R programmers use source
  code control or automated testing.

* Metaprogramming is a double-edged sword. Too many R functions use
  tricks to reduce the amount of typing at the cost of making code that
  is hard to understand and that can fail in unexpected ways.

* Inconsistency is rife across contributed packages, even within base R.
  You are confronted with over 20 years of evolution every time you use R. 
  Learning R can be tough because there are many special cases to remember.

* R is not a particularly fast programming language, and poorly written R code
  can be terribly slow. R is also a profligate user of memory. 