---
title: "Data Structure"
author: "Rochelle Terman"
date: "August 12, 2015"
output: html_document
---

# Classes and Data Structures

To make the best of the R language, you'll need a strong understanding of the basic data types and data structures and how to operate on those.

It is **Very Important** to understand because these are the objects you will manipulate on a day-to-day basis in R. Dealing with object conversions is one of the most common sources of frustration for beginners.

> To understand computations in R, two slogans are helpful:
  - Everything that exists is an object.
  - Everything that happens is a function call.
  
> __John Chambers__

### Main Clasess

`R`'s main atomic classes are:

* character (or a "string" in Python)
* numeric (integer or float)
* integer (just integer)
* logical (booleans)

| Example | Type |
| ------- | ---- |
| "a", "swc" | character |
| 2, 15.5 | numeric | 
| 2 (Must add a `L` at end to denote integer) | integer |
| `TRUE`, `FALSE` | logical |

Like Python, R is a dynamically typed.  There are a few differences in terminology, however, that are pertinent. First, "types" in Python are referred to as "classes"" in R.  Second, R has some different names for the types string, integer, and float --- specifically **character**, **integer** (not different), and **numeric**. Because there is no "float" class in R, users tend to default to the "numeric" class when they want to work with numerical data.

The function for recovering object classes is ```class()```.

```{r}
class(3)
class(3L)
class("Three")
```

But you can also use these helpful functions to understand more about an object

* Structure, `str()`, gives a summary of the object
* Type, `typeof()`, gives storage type of an object - simliar to `class()`
* Length, `length()`, how many elements it contains.
* Attributes, `attributes()`, additional arbitrary metadata.

```{r, eval = FALSE}
x <- "dataset"
typeof(x)
attributes(x)

y <- 1:10
typeof(y)
length(y)
attributes(y)

z <- c(1L, 2L, 3L)
typeof(z)
```


## Data Structures

R's base data structures can be organised by their dimensionality (1d, 2d, or nd) and whether they're homogeneous (all contents must be of the same type) or heterogeneous (the contents can be of different types). This gives rise to the five data types most often used in data analysis: 

|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

Each data structure has its own specifications and behavior.

##### Outline

We'll explain each data structure in this order:

* [Vectors](#vectors) introduces you to atomic vectors and lists, R's 1d 
  data structures.
  
* [Attributes](#attributes) takes a small detour to discuss attributes,
  R's flexible metadata specification. Here you'll learn about factors,
  an important data structure created by setting attributes of an atomic 
  vector.
  
* [Matrices and arrays](#matrices-and-arrays) introduces matrices and arrays,
  data structures for storing 2d and higher dimensional data.
  
* [Data frames](#data-frames) teaches you about the data frame, the most
  important data structure for storing data in R.
  

