---
title: "Data Structures"
author: "Rochelle Terman"
date: "August 12, 2015"
output: html_document
---

# Classes and Data Structures

To make the best of the R language, you'll need a strong understanding of the basic data types and data structures and how to operate on those.

It is **Very Important** to understand because these are the objects you will manipulate on a day-to-day basis in R. Dealing with object conversions is one of the most common sources of frustration for beginners.

> To understand computations in R, two slogans are helpful:
  - Everything that exists is an object.
  - Everything that happens is a function call.
  
> __John Chambers__

### Main Clasess

`R`'s main atomic classes are:

* character (or a "string" in Python)
* numeric (integer or float)
* integer (just integer)
* logical (booleans)

| Example | Type |
| ------- | ---- |
| "a", "swc" | character |
| 2, 15.5 | numeric | 
| 2 (Must add a `L` at end to denote integer) | integer |
| `TRUE`, `FALSE` | logical |

Like Python, R is a dynamically typed.  There are a few differences in terminology, however, that are pertinent. First, "types" in Python are referred to as "classes" in R.  Second, R has some different names for the types string, integer, and float --- specifically **character**, **integer** (not different), and **numeric**. Because there is no "float" class in R, users tend to default to the "numeric" class when they want to work with numerical data.

The function for recovering object classes is ```class()```.

```{r}
class(3)
class(3L)
class("Three")
```

But you can also use these helpful functions to understand more about an object

* Structure, `str()`, gives a summary of the object
* Type, `typeof()`, gives storage type of an object - simliar to `class()`
* Length, `length()`, how many elements it contains.
* Attributes, `attributes()`, additional arbitrary metadata.

```{r, eval = FALSE}
x <- "dataset"
typeof(x)
attributes(x)

y <- 1:10
typeof(y)
length(y)
attributes(y)

z <- c(1L, 2L, 3L)
typeof(z)
```

## Data Structures

R's base data structures can be organised by their dimensionality (1d, 2d, or nd) and whether they're homogeneous (all contents must be of the same type) or heterogeneous (the contents can be of different types). This gives rise to the five data types most often used in data analysis: 

|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

Each data structure has its own specifications and behavior.

##### Outline

We'll explain each data structure in this order:

* [Vectors](#vectors) introduces you to atomic vectors and lists, R's 1d 
  data structures.
  
* [Matrices and arrays](#matrices) introduces matrices,
  data structures for storing 2d and higher dimensional data.
  
* [Attributes](#attributes) takes a small detour to discuss attributes,
  R's flexible metadata specification. Here you'll learn about factors,
  an important data structure created by setting attributes of an atomic 
  vector.
  
* [Data frames](#data-frames) teaches you about the data frame, the most
  important data structure for storing data in R.

## Vectors

A vector is the most common and basic data structure in `R` and is pretty much the workhorse of R. Technically, vectors can be one of two types:

* atomic vectors
* lists

although the term "vector" most commonly refers to the atomic type not lists.

Vectors have three common properties:

* Type, `typeof()`, what it is.
* Length, `length()`, how many elements it contains.
* Attributes, `attributes()`, additional arbitrary metadata.

They differ in the types of their elements: all elements of an atomic vector must be the same type, whereas the elements of a list can have different types.

### Atomic vectors

There are four common types of atomic vectors that I'll discuss in detail: 
* `logical` 
* `integer` 
* `double` (often called `numeric`)
* `character`.

You can create an empty vector with `vector()` (By default the mode is `logical`. You can be more explicit as shown in the examples below.) It is more common to use direct constructors such as `character()`, `numeric()`, etc.

```{r, eval = TRUE}
x <- vector()
# with a length and type
vector("character", length = 10)
character(5) ## character vector of length 5
numeric(5)
logical(5)
```

Atomic vectors are usually created with `c()`, short for combine: \indexc{c()}

```{r, eval = TRUE}
x <- c(1, 2, 3)
x
length(x)
```

`x` is a numeric vector. These are the most common kind. They are numeric objects and are treated as double precision real numbers. To explicitly create integers, add an `L` at the end.

```{r, eval = TRUE}
x1 <- c(1L, 2L, 3L)
```

You can also have logical vectors. 

```{r, eval = TRUE}
y <- c(TRUE, TRUE, FALSE, FALSE)
```

Finally you can have character vectors:

```{r, eval = TRUE}
z <- c("Alec", "Dan", "Rob", "Karthik")
```

**Examine your vector**  

```{r, eval = TRUE}
typeof(z)
length(z)
class(z)
str(z)
```

Question: Do you see a property that's common to all these vectors above?

**Add elements**

```{r, eval = TRUE}
z <- c(z, "Annette")
z
```
More examples of vectors

```{r, eval = TRUE}
x <- c(0.5, 0.7)
x <- c(TRUE, FALSE)
x <- c("a", "b", "c", "d", "e")
x <- 9:100
x <- c(1+0i, 2+4i)
```

You can also create vectors as a sequence of numbers

```{r, eval = TRUE}
series <- 1:10
seq(10)
seq(1, 10, by = 0.1)
```

Atomic vectors are always flat, even if you nest `c()`'s:

```{r}
c(1, c(2, c(3, 4)))
# the same as
c(1, 2, 3, 4)
```

**Types and Tests*

Given a vector, you can determine its type with `typeof()`, or check if it's a specific type with an "is" function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or, more generally, `is.atomic()`. \indexc{typeof()}

```{r}
char_var <- c("harry", "sally")
typeof(char_var)
is.character(char_var)
is.atomic(char_var)

dbl_var <- c(1, 2.5, 4.5)
typeof(dbl_var)
is.double(dbl_var)
is.atomic(dbl_var)
```

NB: `is.vector()` does not test if an object is a vector. Instead it returns `TRUE` only if the object is a vector with no attributes apart from names. Use `is.atomic(x) || is.list(x)` to test if an object is actually a vector.


**Coercion**

All elements of an atomic vector must be the same type, so when you attempt to combine different types they will be __coerced__ to the most flexible type. Types from least to most flexible are: logical, integer, double, and character. \index{coercion}

For example, combining a character and an integer yields a character:

```{r}
str(c("a", 1))
```

**Guess what the following do without running them first**

```{r, eval = FALSE}
xx <- c(1.7, "a") 
xx <- c(TRUE, 2) 
xx <- c("a", TRUE) 
```

Notice that when a logical vector is coerced to an integer or double, `TRUE` becomes 1 and `FALSE` becomes 0. This is very useful in conjunction with `sum()` and `mean()`

```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)

# Total number of TRUEs
sum(x)

# Proportion that are TRUE
mean(x)
```

Coercion often happens automatically. This is called implicit coercion. Most mathematical functions (`+`, `log`, `abs`, etc.) will coerce to a double or integer, and most logical operations (`&`, `|`, `any`, etc) will coerce to a logical. You will usually get a warning message if the coercion might lose information. 

```{r, eval = TRUE}
1 < "2"
"1" > 2
```

You can also coerce vectors explicitly coerce with `as.character()`, `as.double()`, `as.integer()`, or `as.logical()`. Example:

```{r, eval = TRUE}
x <- 0:6
as.numeric(x)
as.logical(x) 
as.character(x)
as.complex(x) 
```

Sometimes coercions, especially nonsensical ones, won’t work.

```{r, eval = TRUE}
x <- c("a", "b", "c")
as.numeric(x)
as.logical(x)
# both don't work
```

## List

Lists are different from atomic vectors because their elements can be of any type. You construct lists by using `list()` instead of `c()`: \index{lists} \index{vectors!lists|see{lists}}

Lists are sometimes called recursive vectors, because a list can contain other lists. This makes them fundamentally different from atomic vectors. 

```{r, eval = TRUE}
x <- list(1, "a", TRUE, 1+4i)
x
```

You can coerce other objects using `as.list()`. You can test for a list with `is.list()`

```{r, eval = TRUE}
x <- 1:10
x <- as.list(x)
is.list(x)
length(x)
```

`c()` will combine several lists into one. If given a combination of atomic vectors and lists, `c()` will coerce the vectors to lists before combining them. Compare the results of `list()` and `c()`:

```{r}
x <- list(list(1, 2), c(3, 4))
y <- c(list(1, 2), c(3, 4))
str(x)
str(y)
```

You can turn a list into an atomic vector with `unlist()`. If the elements of a list have different types, `unlist()` uses the same coercion rules as `c()`.

```{r}
x <- list(list(1, 2), c(3, 4))
x
unlist(x)
```

Lists are used to build up many of the more complicated data structures in R. For example, both data frames (described in [data frames](#data-frames)) and linear models objects (as produced by `lm()`) are lists:

```{r}
is.list(mtcars)

mod <- lm(mpg ~ wt, data = mtcars)
is.list(mod)
```

For this reason, lists are extremely useful inside functions. You can "staple" together lots of different kinds of results into a single object that a function can return.

A list does not print to the console like a vector. Instead, each element of the list starts on a new line.

```{r}
x.vec <- c(1,2,3)
x.list <- list(1,2,3)
x.vec
x.list
```

For lists, elements are indexed by double brackets. Single brackets will still return a(nother) list.


### Exercises

1. What are the four basic types of atomic vector? How does a list differ from an
   atomic vector?

2. Test your knowledge of vector coercion rules by predicting the output of
   the following uses of `c()`:

    ```{r, eval=FALSE}
    c(1, FALSE)
    c("a", 1)
    c(list(1), "a")
    c(TRUE, 1L)
    ```

3.  Why do you need to use `unlist()` to convert a list to an 
    atomic vector? Why doesn't `as.vector()` work? 

4. Why is `1 == "1"` true? Why is `-1 < FALSE` true? Why is `"one" < 2` false?

5. If `x` is a list, what is the class of `x[1]`?  How about `x[[1]]`?

## Matrix

Matrices are a special vector in R. They are not a separate type of object but simply an atomic vector with dimensions added on to it. Matrices have rows and columns. 

```{r, eval = TRUE}
m <- matrix(nrow = 2, ncol = 2)
m
dim(m)
```

Matrices are filled column-wise. 

```{r, eval = TRUE}
m <- matrix(1:6, nrow = 2, ncol = 3)
m
```

Other ways to construct a matrix

```{r, eval = TRUE}
m <- 1:10
dim(m) <- c(2, 5)
m
dim(m) <- c(5, 2)
m
```

You can transpose a matrix with `t()`

```{r}
m <- 1:10
dim(m) <- c(2, 5)
m
t(m)
```

Another way is to bind columns or rows using `cbind()` and `rbind()`.

```{r, eval = TRUE}
x <- 1:3
y <- 10:12
cbind(x, y)
# or 
rbind(x, y)
```

You can also use the `byrow` argument to specify how the matrix is filled. From R's own documentation:

```{r, eval = TRUE, tidy = FALSE}
mdat <- matrix(c(1,2,3, 11,12,13), nrow = 2, ncol = 3, byrow = TRUE,
               dimnames = list(c("row1", "row2"),
                               c("C.1", "C.2", "C.3")))
mdat
```   

Notice that we gave `names` to the dimentions in `mdat`.

```{r}
dimnames(mdat)
rownames(mdat)
colnames(mdat)
```

## Attributes {#attributes}

We just saw that a matrix can have `rownames` and `colnames`. `Names` are a kind of attribute.

We've already seen the following three attributes in action:

* Names, a character vector giving each element a name, described in 
  [names](#vector-names). 

* Dimensions, used to turn vectors into matrices and arrays, 
  described in [matrices and arrays](#matrices-and-arrays).

* Class, used to implement the S3 object system, described in [S3](#s3).
 
Each of these attributes has a specific accessor function to get and set values. When working with these attributes, use `names(x)`, `dim(x)`, and `class(x)`

#### Names {#vector-names}

You can name a vector when you create it: \index{attributes|names}

```{r}
x <- c(a = 1, b = 2, c = 3)
```

You can also modifying an existing vector: 

```{r}
x <- 1:3
names(x) <- c("a", "b", "c")
```

Names don't have to be unique. However, character subsetting, described in [subsetting](#lookup-tables), is the most important reason to use names and it is most useful when the names are unique.

Not all elements of a vector need to have a name. If some names are missing, `names()` will return an empty string for those elements. If all names are missing, `names()` will return `NULL`.

```{r}
y <- c(a = 1, 2, 3)
names(y)

z <- c(1, 2, 3)
names(z)
```

You can create a new vector without names using `unname(x)`, or remove names in place with `names(x) <- NULL`.

### Factors

Factors are special vectors that represent categorical data. Factors can be ordered or unordered and are important when for modelling functions such as `lm()` and `glm()` and also in plot methods.

Factors can only contain pre-defined values. Set allowed values using the `levels()` attribute.

```{r}
x <- factor(c("a", "b", "b", "a"))
x
class(x)
levels(x)

# You can't use values that are not in the levels
x[2] <- "c"

# NB: you can't combine factors
c(factor("a"), factor("b"))
```

Factors are pretty much integers that have labels on them. Underlying it's represented by numbers 1, 2, 3.

```{r}
str(x)
```

They are better than using simple integer labels because factors are what are called self describing. For example, `democrat` and `republican` is more descriptive than `1`s and `2`s. 

Factors are useful when you know the possible values a variable may take, even if you don't see all values in a given dataset. Using a factor instead of a character vector makes it obvious when some groups contain no observations:

```{r}
party_char <- c("democrat", "democrat", "democrat")
party_factor <- factor(party_char, levels = c("democrat", "republican"))

table(party_char)
table(party_factor)
```

Sometimes factors can be left unordered. Example: `democrat`, `republican`.

Other times you might want factors to be ordered (or ranked). Example: `low`, `medium`, `high`. 

```{r}
x <- factor(c("low","medium","high"))
str(x)
is.ordered(x)

x <- ordered(c("low","medium","high"), levels = c("high","medium","low"))
is.ordered(x)
```

While factors look (and often behave) like character vectors, they are actually integers. Be careful when treating them like strings. Some string methods (like `gsub()` and `grepl()`) will coerce factors to strings, while others (like `nchar()`) will throw an error, and still others (like `c()`) will use the underlying integer values. 

```{r}
x <- c("a", "b", "b", "a")
x
is.factor(x)
x <- as.factor(x)
x
c(x, "c")
```

For this reason, it's usually best to explicitly convert factors to character vectors if you need string-like behaviour. In early versions of R, there was a memory advantage to using factors instead of character vectors, but this is no longer the case. \index{factors|)}

Unfortunately, most data loading functions in R automatically convert character vectors to factors. This is suboptimal, because there's no way for those functions to know the set of all possible levels or their optimal order. Instead, use the argument `stringsAsFactors = FALSE` to suppress this behaviour, and then manually convert character vectors to factors using your knowledge of the data.\indexc{stringsAsFactors}

**More attributes**
All R objects can have arbitrary additional attributes, used to store metadata about the object. Attributes can be thought of as a named list (with unique names). Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`. \index{attributes}

```{r}
y <- 1:10
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")
str(attributes(y))
```

### Exercises

1.  What happens to a factor when you modify its levels? 
    
```{r, results = "none"}
f1 <- factor(letters)
levels(f1) <- rev(levels(f1))
```

2.  What does this code do? How do `f2` and `f3` differ from `f1`?

```{r, results = "none"}
f2 <- rev(factor(letters))
f3 <- factor(letters, levels = rev(letters))
```

3.  Here's some code that prints a simple barplot:

```{r}
f <- factor(c("low","medium","high","medium","high","medium"))
table(f)
barplot(table(f))
```

How would you relevel `f` to be in the correct order?

## Data frames {#data-frames}

A data frame is a very important data type in R. It's pretty much the **de facto** data structure for most tabular data and what we use for statistics. Under the hood, a data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list.  This means that a data frame has `names()`, `colnames()`, and `rownames()`, although `names()` and `colnames()` are the same thing. The `length()` of a data frame is the length of the underlying list and so is the same as `ncol()`; `nrow()` gives the number of rows. \index{data frames|(}

### Creation

You create a data frame using `data.frame()`, which takes named vectors as input:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
str(df)
```

Beware `data.frame()`'s default behaviour which turns strings into factors. Use `stringAsFactors = FALSE` to suppress this behaviour: \indexc{stringsAsFactors}

```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE)
str(df)
```

In reality, the most common way to make a data.frame is through `read.csv` or other data file input.

### Testing and coercion

Because a `data.frame` is an S3 class, its type reflects the underlying vector used to build it: the list. To check if an object is a data frame, use `class()` or test explicitly with `is.data.frame()`:

```{r}
typeof(df)
class(df)
is.data.frame(df)
```

You can coerce an object to a data frame with `as.data.frame()`:

* A vector will create a one-column data frame.

* A list will create one column for each element; it's an error if they're 
  not all the same length.
  
* A matrix will create a data frame with the same number of columns and rows as the matrix.

### Combining data frames

You can combine data frames using `cbind()` and `rbind()`: \indexc{cbind()} \indexc{rbind()}

```{r}
cbind(df, data.frame(z = 3:1))
rbind(df, data.frame(x = 10, y = "z"))
```

When combining column-wise, the number of rows must match, but row names are ignored. When combining row-wise, both the number and names of columns must match. Use `plyr::rbind.fill()` to combine data frames that don't have the same columns. 

It's a common mistake to try and create a data frame by `cbind()`ing vectors together. This doesn't work because `cbind()` will create a matrix unless one of the arguments is already a data frame. Instead use `data.frame()` directly:

```{r}
bad <- data.frame(cbind(a = 1:2, b = c("a", "b")))
str(bad)
good <- data.frame(a = 1:2, b = c("a", "b"),
  stringsAsFactors = FALSE)
str(good)
```

The conversion rules for `cbind()` are complicated and best avoided by ensuring all inputs are of the same type.

### Special columns

Since a data frame is a list of vectors, it is possible for a data frame to have a column that is a list: \index{data frames!list in column}

```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)
df
```

Use list columns with caution: many functions that work with data frames assume that all columns are atomic vectors. \index{data frames|)}

### Exercises

1.  What attributes does a data frame possess?

1.  What does `as.matrix()` do when applied to a data frame with 
    columns of different types?

1.  Can you have a data frame with 0 rows? What about 0 columns?

**Other objects**

Missing values are specified with `NA`, which is a logical vector of length 1. `NA` will always be coerced to the correct type if used inside `c()`

```{r}
x <- c(NA, 1)
x
typeof(NA)
typeof(x)
```

`Inf` is infinity. You can have either positive or negative infinity.

```{r, eval = TRUE}
1/0
1/Inf
```

`NaN` means Not a number. It's an undefined value.

```{r, eval = TRUE}
0/0
```

##### Quiz

You can check your answers in [answers](#data-structure-answers).

1. What are the three properties of a vector, other than its contents?

1. What are the four common types of atomic vectors? 

1. What are attributes? How do you get them and set them?

1. How is a list different from an atomic vector? How is a matrix different
   from a data frame?

1. Can a data frame have a column 
   that is a matrix?

## Answers {#data-structure-answers}

1.  The three properties of a vector are type, length, and attributes.
   
1.  The four common types of atomic vector are logical, integer, double 
    (sometimes called numeric), and character. The two rarer types are
    complex and raw.
    
1.  Attributes allow you to associate arbitrary additional metadata to
    any object. You can get and set individual attributes with `attr(x, "y")`
    and `attr(x, "y") <- value`; or get and set all attributes at once with
    `attributes()`.

1.  The elements of a list can be any type (even a list); the elements of 
    an atomic vector are all of the same type. Similarly, every element of 
    a matrix must be the same type; in a data frame, the different columns 
    can have different types.
    
1.  You can make a matrix a column of a data frame with `df$x <- matrix()`, or
    using `I()` when creating a new data frame `data.frame(x = I(matrix()))`.