---
title: "Assignment 5"
author: "YOUR NAME"
date: "THE DATE"
output: html_document
---

## 1. Basics

1.1 Set your working directory to the `PS239T/06_r-basics`

```{r}
#YOUR CODE HERE
```

1.2 Remove all objects from your environment.

```{r}
#YOUR CODE HERE
```

## 2. Vectors and Lists

#### 2.1 Create three vectors:

* a character vector, `titles`, that contain the names of 3 of your favorite movies

* a numeric vector, `year`, that contains the years in which those movies were produced

* a boolean vector `bechdel` that TRUE/FALSE according to whether those movies pass the [bechdel test](http://bechdeltest.com/)

```{r}
# YOUR CODE HERE
```

#### 2.2 Put those three vectors in a list, called `movies`.

```{r}
# YOUR CODE HERE
```

#### 2.3 Index the list `movie` so that it returns (only) the title of the second movie.

```{r}
# YOUR CODE HERE
```

## 3. Factors

3.1 Here's some code that prints a simple barplot:

```{r}
f <- factor(c("low","medium","high","medium","high","medium"))
table(f)
barplot(table(f))
```

#### How would you relevel `f` to be in the correct order?

```{r}
# YOUR CODE HERE



# Test your code
barplot(table(f))
```

#### ## 4. Subsetting and Manipulating Data

Let's work with some real life data! I've prepared for your a basic country-year dataset, with variables:

* Country name
* Country numerical code
* Year
* UN Ideal point
* Polity2 score of regime type (from Polity VI)
* Physical Integrity Rights score (from CIRI dataset)
* Speech Rights score (from CIRI)
* GDP per capita (from World Bank)
* Population (from World Bank)
* Political Terror Scale using Amnesty International reports (from Political Terror Scale project)
* Composite Index of Military Capabilities (Correlates of War)
* Region


Run the code below to set it loaded into R.

```{r}
dat <- read.csv("country-year.csv")
```

#### 4.1 How many rows and columns are in the dataset?

```{r}
#YOUR CODE HERE
```

#### 4.2 Remove the `X` column from the dataet

```{r}
#YOUR CODE HERE
```

#### 4.3 One of the variables is called "gdp.pc.wdi". This stands for "Gross Domestic Product Per Capita, from the World Bank Development Indicators". Change this variable name in the dataset from " "gdp.pc.wdi" to "GDP"

```{r}
#YOUR CODE HERE
```

#### 4.4 Subset the data to include only observations from years 1990-2000 (inclusive). Put the subsetted data in a new variable called `dat.1990.2000`

```{r}
#YOUR CODE HERE
```

#### 4.5 Type the code that calls up the help file on the function `mean()`.

```{r}
#YOUR CODE HERE
```

#### 4.6 Using `mean()` function, tell me the average GDP is in the dataset.

```{r}
#YOUR CODE HERE
```

#### 4.7 You just calculated the average GDP for years 1990-2000. Now calculate the average GDP from 2001 onwards. Tell me how much larger it is (in percentage).

```{r}
#YOUR CODE HERE
```

#### 4.8. What does `summary(dat$polity2[dat$region =="Africa"])` do? Explain below.


#### 4.9  Look up the helpfile for the function `is.na()`. Using this function, replace all the NA values in the `polity2` column of `dat` with 0. 

```{r}
#YOUR CODE HERE
```
