---
title: "Advanced Programming in R"
author: "Rochelle Terman"
date: "Fall 2015"
output: html_document
---

We've already learned about types in R. Now let's discuss some more advanced programming in R - namely:

1. String operations
2. Flow control
3. Functions
4. Loops
5. Vectorized Functions

# 1. Strings

### 1.1 Intro to Operations on Strings

```{r}
firstName <- "Johan"
lastName <- "Gambolputty"
```

Unlike in Python, R does not have a reserved operator for string concatenation such as "+".  Furthermore, using the usual concatenation operator ```c()``` on two or more character strings will not create a single character string, but rather a **vector** of character strings. 

```{r}
firstName <- "Johan"
lastName <- "Gambolputty"

fullName <- c(firstName, lastName)
print(fullName)
length(fullName)
```

In order to combine two or more character strings into one larger character string requires using the ```paste()``` function.  This function takes character strings or vectors and collapses their values into a single character string, with each value separated by a character string selected by the user.

```{r}
firstName <- "Johan"
lastName <- "Gambolputty"

fullName<-paste(firstName, lastName)
print(fullName)

fullName<-paste(firstName, lastName, sep = "+")
print(fullName)

fullName<-paste(firstName, lastName, sep = "___")
print(fullName)
```

As with Python, R can also extract substrings based on the index position of its characters.  There are, however, two critical differences.  First, **index positions in R start at 1**.  This is in contrast to Python, where indexation starts at 0.  Fun huh?

Second, **object subsets using index positions in R contain all the elements in the specified range**.  If some object called ```data``` contains five elements, ```data[2:4]``` will return the elements at the second, third, and fourth positions.  By contrast, the same subset in Python would return the objects at the third and fourth positions (or second and third positions, depending upon whether your index starts at 0 or 1).  

Third, **R does not allow index-based character string subsetting using the object name***, as this functionality is generally reserved for vector subsetting.  Instead, you must use the ```substr()``` function.  Note that this function must receive both the ```start``` and ```stop``` arguments.  So if you want to get all the characters between some index and the end of the string, you must make use of the ```nchar()``` function, which will tell you the length of a character string.

```{r}
firstName <- "Johan"
lastName <- "Gambolputty"
fullName <- paste(firstName, lastName)

# this won't work like in Python
fullName[0]
fullName[1]
fullName[4]
fullName[0:4]
fullName[1:4]

# So use this instead
substr(x = fullName, start = 1, stop = 1)
substr(x = fullName, start = 5, stop = 5)
substr(x = fullName, start = 1, stop = 4)
substr(x = fullName, start = 1, stop = 5)
substr(x = fullName, start = 6, stop = nchar(fullName))
```

### 1.2 Character String Methods

Like Python, R has a number of string methods, though these exist as separate functions. For example:

```{r}
firstName <- "Johan"
lastName <- "Gambolputty"
fullName <- paste(firstName, lastName)

toupper(x = fullName)
tolower(x = fullName)
strsplit(x = fullName, split = " ")
strsplit(x = fullName, split = "a")
gsub(pattern = "G", replacement = "B", x = fullName)
gsub(pattern = "Johan", replacement = "Mike", x = fullName)
gsub(pattern = "johan", replacement = "Mike", x = fullName) # Note the importance of cases!
gsub(pattern = " ", replacement = "", x = fullName) # The same function is used for replacements and stripping
```


# 2. Flow Control

Almost all the conditional operators used in Python also work in R.  The basic loop set up is also very similar, with some small syntax adjustments.  Just as with printing objects, ```if()``` is not a reserved operator in R.  Instead, it is a function whose arguments must be specified inside parentheses.  ```else```, however, is a reserved operator that takes no arguments.  Note that there is no ```elif``` option --- one simply writes ```else if()```.  Whereas operations to be executed after conditional evaluations in Python come after a ```:```, R operations must only be enclosed in curly brackets.  Furthermore, there is no requirement for indentation.  The only thing to keep in mind is that each new operation must be on a separate line.

```{r}
x <- 5
if(x<0){
  print("x is negative")
} 

x <- -5
if(x<0){
  print("x is negative")
}

x <- 5
if(x<0){
  print("x is negative")
} else{
  print("x is positive")
}

x <- 0
if(x<0){
  print("x is negative")
} else if(x==0){
  print("x is zero")
} else{
  print("x is positive")
}
```

R also does some class coercion that makes boolean evaluations harder to break than in Python.  But be careful --- R has a set of special coercions used for fast logical evaluation and subsetting.  Specifically, ```TRUE``` is considered equal to ```1```, while ```FALSE``` is equal to ```0```. The boolean logicals can also be specified as a full word in all caps, or simply as ```T``` or ```F```.

```{r}
1<2
"1"<2
TRUE=="TRUE"
T=="TRUE"
TRUE=="T"
TRUE=="FALSE"
TRUE==0
TRUE==1
TRUE==2
FALSE==0
FALSE==1
FALSE==2
```

# 3. Functions

While functions are defined in Python using the ```def``` reserved operator, R sees functions as just any other object.  Thus, they require explicit assignment to an object.  This is done using the function ```function()```, which creates a function taking the arguments specified in parentheses.  

```{r}
simple.function<-function(x){
  print(x+1)
}
simple.function(x = 2)

less.simple.function<-function(x,y){
  print(x-y+1)
}
less.simple.function(x = 2, y = 10)
```

With respect to returning function output, all the same rules apply as with Python (save for the fact that ```return()``` is a function rather than a reserved operator).  Also be sure to remember that ```return()``` will only process a single object, so multiple items must usually be returned as a list.

```{r}
dumbfun<-function(x){
  return(x)
  print("This will never print :(")
}
dumbfun(x = "something")

dumbfun<-function(x){
  print("This will never print :(")
  return(x)
}
dumbfun(x = "something")

dumbfun<-function(x,y){
  thing1<-x
  thing2<-y
  return(list(thing1, thing2))
}
dumbfun(x = "some text", y = "some data")
dumbfun(x = c(5,10,15), y = "some data")
```

R functions also allow you to set default argument values:

```{r}
less.simple.function<-function(x,y=0){
  print(x-y+1)
}
less.simple.function(x = 2)
less.simple.function(x = 2, y = 10)
```

With respect to specifying arguments, one can either use argument position specifications or argument name specifications.  The latter is strongly preferred, as it is very easy to accidentally specifying incorrect argument values.

```{r}
send<-function(message, recipient, cc=NULL, bcc=NULL){
  print(paste(message, recipient, sep = ", "))
  print(paste("CC:", cc, sep = " "))
  print(paste("BCC:", bcc, sep = " "))
}
send(message = "Hello", recipient = "World", cc = "Rochelle", bcc = "Laura")
send("Hello", "World", "Rochelle", "Laura")
send(message = "Hello", cc = "Rochelle", bcc = "Laura", recipient = "World")
send("Hello", "Rochelle", "Laura", "World")
```

# 4. Loops

Loops in R work basically the same way as in Python, with just a few adjustments.  First, recall that index positions in R start at 1.  Second, ```while()``` and ```for()``` are functions rather than reserved operators, meaning  they must take arguments in parentheses.  Third, just like ```else```, the ```in``` operator *is* reserved and takes no arguments in parentheses.  Fourth, the conditional execution must appear between curly brackets.  Finally, indentation is meaningless, but each new operation must appear on a new line.

```{r}
fruits<-c("apples", "oranges", "pears", "bananas")

# a while loop
i<-1
while(i<=length(fruits)){
  print(fruits[i])
  i<-i+1
}

# a for loop
for(i in 1:length(fruits)){
  print(fruits[i])
}
```

# 5. Vectorized functions

Unlike in Python, while and for loops in R can become very slow.  For this reason, R users generally prefer to use other iteration methods to speed up execution times. 

A common method is the **apply** family of functions. These take a data structure and a function, and applies a function over all the elements in the object.

```{r}
fruit <-c("apple", "orange", "pear", "banana")

# make function that takes in only one element
make.plural <- function(x){
   plural <- paste(x, 's', sep = '')
   return(plural)
}

make.plural('apple')

# apply that function to every element
lapply(fruit, make.plural) # returns a list
sapply(fruit, make.plural) # returns a named vector
```

The multivariate version of `sapply` is `mapply`. Use this if you have a function that takes in 2 or more arguments.