---
title: "Practice"
author: "Rochelle Terman"
date: "October 16, 2015"
output: html_document
---

### setup environment

```{r}
# remove all objects
rm(list=ls())

# set working directory
setwd(dir="~/Dropbox/berkeley/Dissertation/Data\ and\ Analyais/Git\ Repos/PS239T/07_r-data-analysis/")
```

# 1. Loading and summarizing

1.1 Load the gapminder dataset into R.

```{r}
gapminder <- read.csv("data/gapminder-FiveYearData.csv")
head(gapminder)
```

1.2 Tell me the variable names in the dataset

```{r}
names(gapminder)
```

1.3 How many years are covered in the dataset?

```{r}
unique(gapminder$year)
```

1.4 How many unique countries are covered in the dataset?

```{r}
unique(gapminder$country)
```

# 2. Subsetting

2.1 Subset gapminder so that it returns the third column AS ITS OWN A DATAFRAME.

```{r}
gapminder[3]
```

2.2 Subset gapminder so that it returns the third column AS A VECTOR

```{r}
gapminder[[3]]
#OR
gapminder[,3]
```

2.3 Fix each of the following common data frame subsetting errors:

1. Extract observations collected for the year 1957

```{r, eval=FALSE}
gapminder[gapminder$year = 1957,]

# Corrected
gapminder[gapminder$year == 1957,]
```

2. Extract all columns except 1 through to 4

```{r, eval=FALSE}
gapminder[,-1:4]

# Corrected
gapminder[,-c(1:4)]
```

3. Extract the rows where the life expectancy is longer the 80 years

```{r, eval=FALSE}
gapminder[gapminder$lifeExp 80]

# corrected
gapminder[gapminder$lifeExp > 80, ]
```

4. Extract the first row, and the fourth and fifth columns
  (`lifeExp` and `gdpPercap`).

```{r, eval=FALSE}
gapminder[1, 4, 5]

# Corrected
gapminder[1, c(4, 5)]
```

5. Extract rows that contain information for the years 2002
   and 2007

```{r, eval=FALSE}
gapminder[gapminder$year == 2002 | 2007,]

# Corrected
gapminder[gapminder$year == 2002 | gapminder$year == 2007,]
```

# 3. Functions and calculations

3.1. Using `plyr`, tell me the average population in each continent

```{r, eval=FALSE}
# load plyr
library(plyr)

avg.pop.by.region <- ddply(.data = gapminder, .variables = .(continent), .fun = summarize, avg.pop = mean(pop, na.rm = T))

avg.pop.by.region
```

3.2 Merge this data with gapminder, so that you have a new column with the continent population mean.

```{r, eval=FALSE}
# load plyr
merged <- merge(gapminder, avg.pop.by.region, by = c("continent"), all.x = T)

head(merged)
```

3.4 Which observation has the highest population? 

```{r, eval=FALSE}
# two ways to do this.
# 1. using 'max' function.
gapminder[gapminder$pop == max(gapminder$pop),]

# using "arrange"

sorted.by.pop <- arrange(gapminder, desc(pop))
sorted.by.pop[1,]
```

3.5 Advanced: Add a new column, `pop.proportion` that gives the population in terms of proportion of world population total (in that year).

```{r, eval=FALSE}
# first, calculate the total world population for each year. We can use the function "mutate" in plur to automatically add this column in with gapminder.
gapminder.world.pop <- ddply(.data = gapminder, .variables = .(year), .fun = mutate, world.pop = sum(pop, na.rm = T))
# check it out
head(gapminder.world.pop)
# now make new column dividing pop by world.pop
gapminder.world.pop$pop.proportion <- gapminder.world.pop$pop / gapminder.world.pop$world.pop
# check it out.
head(gapminder.world.pop)
```
