---
title: "Pre-processing"
author: "Rochelle Terman"
date: "June 24, 2015"
output: html_document
---

### Required Packages

First let's load our required packages for this unit.

```{r}
setwd("~/Dropbox/berkeley/Dissertation/Data and Analyais/Git Repos/PS239T/text-analysis")
rm(list=ls())
library(tm) # Framework for text mining
library(RTextTools) # a machine learning package for text classification written in R
library(mallet) # a wrapper around the Java machine learning tool MALLET
library(qdap) # quantiative discourse analysis of transcripts
library(qdapDictionaries)
library(dplyr) # data preparation and pipes $>$
library(ggplot2) # for plotting discourse data
library(SnowballC) # for stemming
library(data.table) # for easier data manipulation
library(scales) # to help us plot
library(matrixStats) # for statistics
```

## 1. Prepare a Corpus

A corpus is a collection of texts, usually stored electronically, and from which we perform our analysis. A corpus might be a collection of news articles from Reuters or the published works of Shakespeare. Within each corpus we will have separate articles, stories, volumes, each treated as a separate entity or record. Each unit is called a "document"

Documents come in a variety of formats, but plain text is best.

### 1.1 Corpus Sources and Readers

The `tm` package supports a variety of sources and formats. 

```{r}
getSources()
getReaders()
```

Here we'll be going over two main options to input a corpus. 

Option 1) We can read a corpus from a directory that contains text files, each document a different file.
```{r}
docs <- Corpus(DirSource("Data/MachText")) 
docs
```

Option 2) We can read from a csv of documents, with each row being a document, and columns for text and metadata (information about each document). This is the easiest option if you have metadata.

```{r}
docs.df <-read.csv("Data/mach.csv", header=TRUE) #read in CSV file
docs <- Corpus(VectorSource(docs.df$text))
docs
```

Once we have the corpus, we can inspect the documents using inspect()

```{r}
# see the 16th document
inspect(docs[16])
```

And see the text using the `as.chracter`

```{r}
 # see content for 16th document
as.character(docs[[16]])
```

### 1.2 Preprocessing

Many text analysis applications follow a similar 'recipe' for preprecessing, involving:

1. Tokenizing the text to unigrams (or bigrams, or trigrams)
2. Converting all characters to lowercase
3. Removing punctuation
4. Removing numbers
5. Removing Stop Words, inclugind custom stop words
6. "Stemming" words, or lemmitization. There are several stemming alogrithms. Porter is the most popular.
7. Removing Sparse Terms
8. Creating a Document-Term Matrix

See what transformations are available TM package

```{r}
getTransformations()
```

The function `tm_map()` is used to apply one of these transformations across all documents. 

We can also use regular R functions for custom transformations using the `content_transformer()` function.

For example, we might want to replace “/”, used sometimes to separate alternative words, with a space. This will avoid the two words being run into one string of characters through the transformations. We might also replace “@” and “|” with a space, for the same reason
```{r}
# note: won't work on past versions of R
toSpace <- content_transformer(function(x, pattern) gsub(pattern, " ", x))
docs <- tm_map(docs, toSpace, "/")
docs <- tm_map(docs, toSpace, "@")
docs <- tm_map(docs, toSpace,"\\|")
as.character(docs[[16]])
```

We can basic transformations using the build in `tm_map` tools.

```{r}
docs <- tm_map(docs, content_transformer(tolower)) # convert all text to lower case
as.character(docs[[16]])
docs <- tm_map(docs, removePunctuation) # remove Puncturation
as.character(docs[[16]])
docs <- tm_map(docs, removeNumbers) # remove Numbers
as.character(docs[[16]])
docs <- tm_map(docs, removeWords, stopwords("english")) # remove common words
stopwords("english") # check out what was removed
as.character(docs[[16]])
docs <- tm_map(docs, removeWords, c("prince")) # remove own stop words
as.character(docs[[16]])
docs <- tm_map(docs, stripWhitespace) # strip white space
as.character(docs[[16]])
docs <- tm_map(docs, stemDocument) # stem the document
as.character(docs[[16]])
```

### 1.3 Creating a DTM

A document term matrix is simply a matrix with documents as the rows and terms as the columns and a count of the frequency of words as the cells of the matrix. We use `DocumentTermMatrix()` to create the matrix:
```{r}
dtm <- DocumentTermMatrix(docs)
dtm
# how many documents? how many terms?
dim(dtm)
# take a quick look
inspect(dtm[1:5,1000:1005])
```

Note: we can use RTextTools to go directly from text vector to DTM, which is probably easier. If you read your corpus using Option 1 above (Section 1.1) you must first convert your corpus into a dataframe

```{r}

# Read corpus directly from dir and convert to data.frame.
docs <- Corpus(DirSource("Data/MachText"))
docs.df<-data.frame(text=unlist(sapply(docs, `[`, "content")), 
    stringsAsFactors=F)
```

If you read in your corpus as a CSV, you should already have a dataframe (in this case, `docs.df`) in your environment.

```{r}
# Read corpus from csv
docs.df <- read.csv("Data/mach.csv")
```

Now we can make out DTM in one step.

```{r}
# Convert to DTM
dtm <- create_matrix(docs.df$text, language="english", removeNumbers=TRUE,
                       stemWords=TRUE, toLower = TRUE, removePunctuation = TRUE)
```

### 1.4 Exploring the DTM

We can obtain the term frequencies as a vector by converting the document term matrix into a matrix and summing the column counts:

```{r}
 # how many terms?
freq <- colSums(as.matrix(dtm))
length(freq)
```

By ordering the frequencies we can list the most frequent terms and the least frequent terms:

```{r}
# order
ord <- order(freq)

# Least frequent terms
freq[head(ord)]

# most frequent
freq[tail(ord)]

# frequency of frenquencies
head(table(freq),15)
tail(table(freq),15)

# plot
plot(table(freq))
```

Exploring word frequences

```{r}
# Have a look at common words
findFreqTerms(dtm, lowfreq=100) # words that appear at least 100 times

# Which words correlate with "war"?
findAssocs(dtm, "war", 0.3)

# plot
freq <- sort(colSums(as.matrix(dtm)),decreasing=TRUE)
head(freq)

wf <- data.frame(word=names(freq), freq=freq)
head(wf)

library(ggplot2)
subset(wf, freq>50) %>%
  ggplot (aes(word, freq)) +
  geom_bar (stat ="identity") +
  theme(axis.text.x=element_text(angle=45,hjust=1))

# wordcoulds!
library(wordcloud)
set.seed(123)
wordcloud(names(freq), freq, max.words=100, colors=brewer.pal(6,"Dark2"))
```
We can remove sparse terms and thus inrease efficency

```{r}
dtm.s <- removeSparseTerms(dtm,.99)
dtm # 2365 terms
dtm.s # 1326 terms
```

### 1.5 Exporting the DTM

We can convert a DTM to a matrix or data.frame in order to write to a csv, add meta data, etc.

```{r}
# coerce into dataframe
dtm <- as.data.frame(as.matrix(dtm))
names(docs)

# add fake column for section
dtm$section[1:100] <- "Section 1"
dtm$section[101:188] <- "Section 2"
dtm$section <- as.factor(dtm$section)

# check to see if they're the same number of documents per author
summary(dtm$section)
```
