Sentiment Analysis
========================================================

The `qdap`(Rinker, 2014) package provides an extensive suite of functions to support the quantitative analysis of text. 

The package was originally intended to analyze transcript data. So the data must be in a particular format, namely a tabular spreadsheet with a `text` variable and some kind of `speaker` or metadata variable. The `text` variable must be in the form of sentences, with each sentence ending with punctuation: `.` , `?`, `!`.

`qdap` has an [extensive vignette](https://dl.dropboxusercontent.com/u/61803503/packages/qdap_vignette.html) that will help guide you through its requirements and functionality.

## 1. Comparing Songs on the Thriller Album

Let's start with a simple example using lyrics from Michael Jackson's Thriller album.

```{r}
# load data
thriller <- read.csv("Data/thriller.csv")

# see a truncated version of the data
head(truncdf(thriller),10)

# split the data into sentences (or, in this case, lines in the songs)
thril.split <- sentSplit(thriller, "Lyrics")
head(truncdf(thril.split),10)
```

The `qdap` package is great for using dictionary methods to analyze text. One of the most popular of these menthods is sentiment analysis, which calculates how "positive" or "negative" text is.

In `qdap`, we analyze sentiment using the `polarity` function.

```{r}
# calculate polarity
(poldat <- with(thril.split, polarity(Lyrics, Song)))

# have a peak of each line
counts(poldat)[1:10,]

# plot
plot(poldat)

# we can get just the second plot this way:
x <- plot(poldat)
x$p2
```

## 2. Visualizing Plot using Sentiments

`qdap` also has several datasets pre-formatted to help you explore. Let's conduct a plot analysis of Romeo and Juliet using sentiments.

```{r}
# take a look at the data
head(truncdf(rajSPLIT),10)

# put that into a data.table
raj.dat <- data.table(rajSPLIT)
```

Now that we have the data, we can add columns for cumulative word counts and polarity scores. This will allow us to capture sentiments over "time." 

```{r}
# add word counts
raj.dat[, wc := wc(dialogue,missing=0)]

# Add cumulative word count and percent complete as proxy for progression
raj.dat[, cumsum := cumsum(wc)]
raj.dat[, pct.complete := raj.dat$cumsum / sum(raj.dat$wc)]
raj.dat[, pct.complete.100 := pct.complete * 100]

# calculate polarity
poldat.raj <- with(rajSPLIT, polarity(dialogue, act, constrain = TRUE))
counts(poldat.raj)[1:10,]
polcount.raj <- na.omit(counts(poldat.raj)$polarity) # get vector of just the polarity scores

# put it into a data frame
len <- length(polcount.raj)
pol.df.raj <- data.frame(polarity = polcount.raj, Time=1:len)
```

Let's plot it.

```{r}
## Calculate background rectangles for RAJ
ends <- cumsum(rle(counts(poldat.raj)$act)$lengths)
ends
starts <- c(1, head(ends + 1, -1))
starts
rects <- data.frame(xstart = starts, xend = ends + 1,
                    Act = c("I", "II", "III", "IV", "V"))

# Plot RAJ
ggplot() + theme_bw() +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend,
                              ymin = -Inf, ymax = Inf, fill = Act), alpha = 0.17) +
  geom_line(data = pol.df.raj, aes(y=polarity, x = Time), size=1, color = "grey60") +
  geom_smooth(data = pol.df.raj, aes(y=polarity, x = Time), color="royalblue", fill="lightgray", size=1.4) +
  geom_hline(y=mean(polcount.raj), color="grey30", size=1, alpha=.3, linetype=2) +
  annotate("text", x = mean(ends[1:2]), y = mean(polcount.raj), color="grey30",
           label = "Average Polarity", vjust = .3, size=3) +
  ylab("Average Polarity") + xlab("Duration") +
  scale_x_continuous(expand = c(0,0)) +
  geom_text(data=rects, aes(x=(xstart + xend)/2, y=-.04,
                            label=paste("Act", Act)), size=3) +
  guides(fill=FALSE) +
  scale_fill_brewer(palette="Set1")


# Zoom in on that smoothed line:

ggplot() + theme_bw() +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend,
                              ymin = -Inf, ymax = Inf, fill = Act), alpha = 0.17) +
  #geom_line(data = pol.df.raj, aes(y=polarity, x = Time), size=1, color = "grey60") +
  geom_smooth(data = pol.df.raj, aes(y=polarity, x = Time), color="royalblue", fill="lightgray", size=1.4) +
  geom_hline(y=mean(polcount.raj), color="grey30", size=1, alpha=.3, linetype=2) +
  annotate("text", x = mean(ends[1:2]), y = mean(polcount.raj), color="grey30",
           label = "Average Polarity", vjust = .3, size=3) +
  ylab("Average Polarity") + xlab("Duration") +
  scale_x_continuous(expand = c(0,0)) +
  geom_text(data=rects, aes(x=(xstart + xend)/2, y=-.04,
                            label=paste("Act", Act)), size=3) +
  guides(fill=FALSE) +
  scale_fill_brewer(palette="Set1")


```

## 3. Comparing Plots

Let's use the technique we just applied above , let's compare the plots between Romeo and Julet and Hamlet.
```{r}

# get Hamlet data
ham.dat <- data.table(hamlet)
# add word counts
ham.dat[, wc := wc(dialogue,missing=0)]
# Add cumulative word count and percent complete as proxy for progression
ham.dat[, cumsum := cumsum(wc)]
ham.dat[, pct.complete := ham.dat$cumsum / sum(ham.dat$wc)]
ham.dat[, pct.complete.100 := pct.complete * 100]

# calculate polarity
poldat.ham <- with(hamlet, polarity(dialogue, act, constrain = TRUE))
truncdf(counts(poldat.ham))[1:10,]
polcount.ham <- na.omit(counts(poldat.ham)$polarity) # get vector of just the polarity scores

# put it into a data frame
len.ham <- length(polcount.ham)
pol.df.ham <- data.frame(polarity = polcount.ham, Time=1:len.ham)

# plot HAMLET + RAJ
ggplot() +
  geom_smooth(data = pol.df.raj, aes(y=polarity, x = Time, color="Romeo and Juliet", fill="Romeo and Juliet"), size=1.4) +
  geom_smooth(data = pol.df.ham, aes(y=polarity, x = Time, color="Hamlet", fill="Hamlet"), size=1.4) +
  geom_hline(y=mean(polcount.raj), color="gray30", size=1, alpha=.3, linetype=2) +
  annotate("text", x = mean(ends[1:2]), y = mean(polcount.raj), color="grey30",
           label = "Average Polarity - RAJ", vjust = .3, size=3) +
  geom_hline(y=mean(polcount.ham), color="gray30", size=1, alpha=.3, linetype=2) +
  annotate("text", x = mean(ends[1:2]), y = mean(polcount.ham), color="grey30",
           label = "Average Polarity - HAM", vjust = .3, size=3) +
  ylab("Average Polarity") + xlab("Duration") +
  scale_x_continuous(expand = c(0,0)) +
  geom_text(data=rects, aes(x=(xstart + xend)/2, y=-.04,
                           label=paste("Act", Act)), size=3) +
  guides(fill=F) +
  scale_fill_brewer(palette="Set1") +
  scale_colour_manual('Legend', breaks = c('Romeo and Juliet', 'Hamlet'),
                      values = c('red', 'royalblue'))
```

## 4. Extra: More on QDAP

Besides polarity, `qdap` has other dictionaries to measure language formality, diversity, readability, etc. It also carries a range of tools to analyze and visualize discourse data, word usage, character usage, parts of speech, etc. 

```{r}
# gantt plots
dat <- gantt(rajSPLIT$dialogue, rajSPLIT$person)
head(dat, 12)
plot(dat)

# by grouping
dat2 <- with(rajSPLIT, gantt(dialogue, list(fam.aff, sex)))
head(dat2, 12)
plot(dat2)

# ToT plots
tot_plot(mraja1, "dialogue", grouping.var = c("fam.aff","sex"), tot = FALSE)
```